Strict Pure Functional Programming and Automatic Parallelization\\ (New-project-in-progress Paper)

### 2016-03-28 pubs/tfp16/paper.tex
incremental work on MHS writeup

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New-project-in-progress Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.
\\\ \\
\emph{Preliminatry paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the workshop.

  \item While this does not qualify as a student paper, students
    (undergraduate!)  have contributed considerably to this effort, and if
    accepted we plan to have one of these students present this paper.

  \item Hopefully in the spirit of a student-friendly workshop, for this draft
    version we have chosen a more informal, conversational style of
    presentation than necessarily appropriate for a final paper, have
    highlighted the major student contributions and involvement, and included
    perhaps gratuitous detail about process.
  \end{enumerate}

\keywords{strict pure functional programming, automatic parallelization, Haskell, STG}

@@

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task is prescribed access privileges (read, write,
read/write, etc.)  and coherency requirements for each region to which it will
have access.  Other than registered access to regions, tasks must not access
non-constant global data.  In serial execution these access requirements are
notionally superfluous, but in parallel execution the runtime can dynamically
calculate the data dependency graph and relax the programmatically defined
serial ordering of task execution.  Very high parallel efficiency has been
demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
LANL the Legion system is being strongly considered for a new large-scale C++
code effort.\footnote{It seems significant that the PI of the Legion
  project was previously a co-developer of a strict, arguably pure functional
  language implementation~\cite{AikenFL,FLreport89}, in collaboration with
John Backus as a refinement of his FP language~\cite{Backus:1978}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84}
or Johnsson's general attribute grammars~\cite{Johnsson87}.

@@

A secondary, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of
non-strictness.\footnote{The starting experience of our student interns to
  date has been normalized by having them complete Hutton's \emph{Programming in Haskell} textbook
as baseline preparation.~\cite{Hutton-book}}

@@

\paragraph{Fancy Pattern Matching} -\\

{\color{red}\dots}

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New-project-in-progress Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Strictness constrains
the degree of potential parallelism, that is, we are not exploring speculative
evaluation.

\emph{Preliminary paper note 1: We are reporting work in progress.  At the
  time of writing we have a complete serial implementation, a primitive
  proof-of-concept parallel implementation, and are currently developing a
  more realistic parallel runtime.  We expect to have preliminary results from
  our new parallel implementation by the time of the symposium.}

\emph{Note 2: While this does not qualify as a student paper, students
  (undergraduate!)  have contributed considerably to this effort, and if
  accepted we plan to have one of these students present this paper.}

\emph{Note 3: Hopefully in the spirit of a student-friendly workshop, in this
  draft we have chosen a more conversational style of presentation than
  necessarily appropriate for a final paper, and have also highlighted the
  major student contributions, and perhaps gratuitous detail about process.}

\keywords{strict pure functional programming, automatic parallelization}

@@

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task is prescribed access privileges (read, write,
read/write, etc.)  and coherency requirements for each region to which it will
have access.  Other than registered access to regions tasks must not access
non-constant global data.  In serial execution these access requirements are
notionally superfluous, but in parallel execution the runtime can dynamically
calculate the data dependency graph and relax the programmatically defined
serial ordering of task execution.  Very high parallel efficiency has been
demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
LANL the Legion system is being strongly considered for a new large-scale C++
code effort.\footnote{It seems significant that the PI of the Legion
  project was previously a co-developer of a strict, arguably pure functional
  language implementation~\cite{AikenFL,FLreport89}, in conjunction with
John Backus as a refinement of his FP language~\cite{Backus:1978}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and laziness is more of a nuisance--even a serious
hindrance--than a convenience, much less ever algorithmically essential in the
sense of Bird's \emph{repmin}~\cite{Bird84} or Johnsson's general attribute
grammars~\cite{Johnsson87}.

@@

@@

\paragraph{Nested Pattern Matching} -\\

STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments (with multiple non-anonymous definitions allowed to exhaust
the pattern combinations).  Patterns in all contexts (functions and case
expressions) may be nested arbitrarily deep.  These two features greatly
simplify the programmer's job.
